A Bayesian networks approach to infer social changes in burials in northeastern Taiwan during the European colonization period

### 2021-05-27 analysis/paper/paper.Rmd
edit the main text based on results of 30000 iterations with more chains

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, the ritual element represented by pots (ritual-homophily) has a significant effect on the formation of relations between burials, while the wealth rank represented by burial values (wealth-homophily) has a negative effect but not significant since its confidence intervals include zero. This demonstrates that burials with ritual pottery tend to form relations, but burials in the same wealth level tend not to. Despite positive mean for some covariates, such as age (age-homophily) and sex (sex-homophily), they do not show a significant tendency. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not reflect similarity in burial treatment. For the endogenous network effects, transitivity presents a significant positive effect, while centralization demonstrates a negative effect. The high positive value for transitivity suggests a tendency of burials with similar burial goods to be clustered as connected communities, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative centralization shows there is a tendency toward decentralization that reflects most burials having a similar number of ties without any prominent burials. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

<!--posterior stats for for post-E and their comparison-->
For the post-European network model, the nodal covariate of wealth rank shows significant positive effects. This indicates the burials in the same wealth level tend to have relations. Despite negative effects for the ritual and sex variables, they are not significant. Also, there are no significant effects for the variables of age and physical distance. Similar to the pre-European network, transitivity demonstrates a significant positive effect, but much weaker than the effect of the pre-European network. In contrast, centralization has a significantly higher positive effect than the effect of the pre-European network. This means there is a tendency toward centralization, reflecting a limited number of burials having many more ties than others. This implies that the presence of better access to trade goods and the behavior of wealth accumulation and display in burial events. In general, the post-European network model has a smaller transitivity effect and a positive centralization effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence.

@@

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestige goods. Our results indicate that the major factor contributing to burial networks for the pre-European network was ritual, and for the post-European network was wealth. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors were important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual could be a status indicator at the time before many foreign goods were introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences, corresponding with the characteristics of corporate/network strategies. Geographical proximity of burials, implying kinship association, has no effect on sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Demographic information, such as age and sex, do not have any effects on network formation for both networks. We also note that no effects could be due to the unavailability of age (`r age_miss` out of `r pre_and_post` burials) and sex (`r sex_miss` out of `r pre_and_post` burials) values in our sample.

@@

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
Our case study examined social changes in a European pericolonial context in northeastern Taiwan. We tested a hypothesis of changes in burial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our model that the relationship between burials changed after the European colonization period in the 17th century. Before the arrival of Europeans, the burial network has a tendency of more clustered subgroups with pottery as ritual practice as the key formation mechanism. After European arrival, the network has a tendency of centralization relative to the rarity of goods. We interpret the changes in formation mechanisms of networks, observed as a change from ties formed by ritual to wealth, as an increase in wealth accumulation behaviors, stimulated by the European presence and its associated long-distance trade network. This aligns with changes in horizontal hierarchy represented by corporate-network societies, that demonstrates different degrees of social inequality in relation to different strategies for achieving power [@Feinman2000]. Using burial data with historical documents, we can better understand indirect European impacts in understudied pericolonial contexts of East Asia [e.g. @Acabado2017; @Trabert2017].

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the data used in this paper. We thank the excavation team of Kiwulan led by Yu-pei Chen for their post-excavation analysis and detailed documentation about burials that enable this study. We also thank Alberto Caimo and Steven M. Goodreau for the discussion regarding the methods at the earliest stage of analysis. This work was supported in part by the travel grant from the Department of Anthropology at the University of Washington, the predissertation summer travel grant from the Henry Luce Foundation/ACLS Program in China Studies, and the Doctoral Dissertation Fellowship from the Chiang Ching-kuo Foundation for International Scholarly Exchange (Project #DF009-A-18). We thank Ben Fitzhugh and Peter Lape for their insightful comments on an early draft. We also acknowledge comments from anonymous reviewers that have greatly improved this manuscript.

>>>>

<!--posterior stats for pre-E-->
We examine the estimates from the posterior distributions to compare their differences in structure of the simulated networks (Table \@ref(tab:table-posterior-stats); Figure \@ref(fig:figure-posterior-distribution)). For nodal covariates in the pre-European model, the values of ritual element represented by pots (ritual-homophily) are mostly greater than zero that suggests a positive effect on the formation of relations between burials, while the values of wealth rank represented by burial values (wealth-homophily) are mostly negative that indicates it has less influence on the tie formation. This demonstrates that burials with ritual pottery tend to form relations, but burials in the same wealth level tend not to. For other covariates with values mostly around zero, such as age (age-homophily) and sex (sex-homophily), they do not show any clear tendency. Similarly, the dyadic covariate, physical distance, shows no significant effect, indicating that physical proximity between burials does not reflect similarity in burial treatment. For the endogenous network effects, transitivity presents a significant positive effect, while centralization demonstrates a negative effect. The high positive value for transitivity suggests a tendency of burials with similar burial goods to be clustered as connected communities, indicative of the presence of multiple corporate groups sharing burial goods in common. In contrast, the strong negative centralization shows there is a tendency toward decentralization that reflects most burials having a similar number of ties without any prominent burials. This might imply that individuals have equal access to trade goods in terms of the flow of goods.

@@

<!--posterior stats for for post-E and their comparison-->
For the post-European network model, two nodal covariates, wealth rank and age, both show tendencies towards positive values. This indicates the burials in the same wealth level and age groups tend to form relations. However, there is no clear pattern for ritual and sex variables since most values around zero. Also, there are no significant effects for the variables of physical distance. Similar to the pre-European network, transitivity demonstrates a significant positive effect, but weaker than the effect of the pre-European network. In contrast, centralization has a significantly higher positive effect than the effect of the pre-European network. This means there is a tendency toward centralization, reflecting a limited number of burials having many more ties than others. This implies that the presence of better access to trade goods and the behavior of wealth accumulation and display in burial events. In general, the post-European network model has a smaller transitivity effect and a positive centralization effect than the pre-European network model. This may suggest a reduced tendency toward clustering but high tendency toward centralization after the European presence.

@@

<!--archaeological implication based on the results-->
Network covariates give insights into the socio-cultural factors that were associated with the acquisition and distribution of foreign prestige goods. Our results indicate that the potential factor contributing to burial networks for the pre-European network was ritual, and for the post-European network was wealth and age groups. We interpret these results as illustrating different social and economic mechanisms influencing relations between burials for the two periods. Before European arrival, ritual behaviors could be important factors that structured the interconnections of prestige goods in the burial contexts. This might imply that ritual as a status indicator at the time before many foreign goods were introduced to Indigenous societies. After the European presence was established, status indicators shifted from ritual to wealth differences, corresponding with the characteristics of corporate/network strategies. Geographical proximity of burials, implying kinship association, has no effect on sharing similar prestige items between burials for both phases. This may suggest that the use of prestige items is more individual-oriented instead of kinship-based. Another demographic information, sex, do not have any effects on network formation for both networks. We also note that their weak effects may be impacted by the unavailability of age (`r age_miss` out of `r pre_and_post` burials) and sex (`r sex_miss` out of `r pre_and_post` burials) values in our sample.

@@

<!-- summary of the research by answering questions raised in Intro, and anthropological contribution -->
Our case study examined social changes in a European pericolonial context in northeastern Taiwan. We tested a hypothesis of changes in burial networks at Kiwulan with the evaluation of both endogenous and exogenous network effects. The results support our model that the relationship between burials changed after the European colonization period in the 17th century. Before the arrival of Europeans, the burial network has a tendency of more clustered subgroups with pottery as ritual practice as the key formation mechanism. After European arrival, the network has a tendency of centralization relative to the rarity of goods. We interpret the changes in formation mechanisms of networks, observed as a change from tie formation dominated by ritual to wealth, as an increase in wealth accumulation behaviors, stimulated by the European presence and its associated long-distance trade network. This aligns with changes in horizontal hierarchy represented by corporate-network societies, that demonstrates different degrees of social inequality in relation to different strategies for achieving power [@Feinman2000]. Using burial data with historical documents, we can better understand indirect European impacts in understudied pericolonial contexts of East Asia [e.g. @Acabado2017; @Trabert2017].

@@

We would like to thank the Yilan County Cultural Affairs Bureau in Taiwan for permitting access to the data used in this paper. We thank the excavation team of Kiwulan led by Yu-pei Chen for their post-excavation analysis and detailed documentation about burials that enable this study. We also thank Alberto Caimo and Steven M. Goodreau for the discussion regarding the methods at the earliest stage of analysis. This work was supported in part by the travel grant from the Department of Anthropology at the University of Washington, the predissertation summer travel grant from the Henry Luce Foundation/ACLS Program in China Studies, and the Doctoral Dissertation Fellowship from the Chiang Ching-kuo Foundation for International Scholarly Exchange (Project #DF009-A-18). We thank Ben Fitzhugh, Peter Lape, Erik Gjesfjeld, Matthew Peeples for their insightful comments on an early draft. We also acknowledge comments from anonymous reviewers that have greatly improved this manuscript.
